Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-03 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Remove or rewrite parts in bold face until results

Anyone who analyzes person-data may run into a missing data problem.
Missing data is not only ubiquitous, but treating it can also be
tedious. If a dataset contains just one incomplete observation,
statistical inferences are undefined and will not produce any results.
To circumvent this, many statistical packages employ list-wise deletion
by default (i.e., ignoring incomplete observations). Unfortunately, this
\emph{ad hoc} solution may yield wildly invalid results \citep{buur18}.
An alternative is to \emph{impute} (i.e., fill in) the missing values in
the incomplete observations. Subsequently, statistical inferences can be
performed on the completed dataset. By repeating this process several
times, a distribution of plausible results may be obtained, which
reflects the uncertainty in the data due to missingness. This technique
is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven
to be a powerful technique to yield unbiased and confidence valid
estimates of the true---but missing---data inference under many
circumstances \citep{buur18}.

Figure \ref{fig:diagram} provides an overview of the steps involved with
MI---from incomplete data, to \(M\) multiply imputed datasets, to \(M\)
estimated quantities of interest \(\hat{Q}\)s, to a single pooled
estimate \(\bar{Q}\). Missing data in dataset \(y\) is imputed \(M\)
times. The imputed data \(y_{imp}\) is combined with the observed data
\(y_{obs}\) to create \(M\) completed datasets. On each completed
dataset, the analysis of scientific interest is performed. The quantity
of scientific interest (e.g., a regression coefficient) is denoted with
\(Q\). Since \(Q\) is estimated on each completed dataset, \(M\)
separate \(\hat{Q}\)-values are obtained. These \(M\) values are
combined into a single pooled estimate \(\bar{Q}\).

@@

\caption{Scheme of the main steps in multiple imputation.}\label{fig:diagram}
\end{figure}

A popular method to obtain imputations is to use the `Multiple
Imputation by Chained Equations' algorithm, shorthand
`MICE'\citep{mice}. MICE is an iterative algorithmic procedure to draw
imputations from the posterior predictive distribution of the missing
values. This introduces a potential threat to the validity of the
imputations: What if the algorithm has not converged? Are the
implications then to be trusted? And can we rely on the inference
obtained on the completed data? These are all open questions, because
the convergence properties of iterative imputation algorithms have not
been systematically studied \citep{buur18}. Moreover, there is no
scientific consensus on how to evaluate convergence of MI algorithms
\citep{taka17}. Some default MICE techniques (e.g., `predictive mean
modeling') might not yield converged states at all \citep{murr18}.
Therefore, algorithmic convergence should be monitored carefully.

Currently, the recommended practice for evaluating convergence in
iterative imputation algorithms is to visually inspect imputations for
signs of non-convergence. \textbf{Add that this refers to chain means
and variances already? It's insufficient that these are univariate, so
multivariate state space of the algorithm may not be converged when
univariates are ok. And \citet{buur18}`s suggestion to track
user-defined statistics (e.g., regression coefficient) may be somewhat
advanced for empirical researchers. Moreover, this is not
model-independent, i.e., only apply to one substantive model, while one
of the advantages of MI is that missing data problem and scientific
problem are split. Then we can say 'we propose a novel\ldots{}' and
integrate this: Note that usually we only evaluate the convergence of
univariate scalar summaries (e.g., chain means or variances). With these
we cannot diagnose convergence of multivariable statistics (i.e.,
relations between scalar summaries). Van Buuren \citeyearpar[\(\S\)
4.5.2]{buur18} proposed to implement multivariable evaluation of the
MICE algorithm through eigenvalue decomposition building on the work of
\citet{mack03}. Eigenvalues of a covariance matrix are a measure of the
data's covariance.}.

This method is insufficient on two counts: 1) it may be challenging to
the untrained eye, and 2) only severely pathological cases of
non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}.
Therefore, a quantitative, diagnostic evaluation of convergence would be
preferred---yet not straightforward. Iterative imputation algorithms
such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC
methods, convergence is not from a scalar to a point, but from one
distribution to another. The values generated by the algorithm (e.g.,
imputed values) will vary even after convergence. Since MCMC algorithms
do not reach a unique point at which convergence is established,
diagnostics we may only identify signs of non-convergence
\citep{hoff09}. Several of such diagnostics exist, but it is not known
whether these are appropriate for iterative imputation algorithms.

In this paper, we investigate how non-convergence in iterative
imputation algorithms may be diagnosed, how well these methods perform,
and at which point convergence may safely be assumed. For reasons of
brevity, we only focus on the iterative imputation algorithm implemented
in the popular \texttt{mice} package \citep{mice} in \texttt{R}
\citep{R}. The convergence properties of the MICE algorithm are
investigated through model-based simulation. The results of this
simulation study are guidelines for assessing convergence of MI
algorithms, which will aid applied researchers in drawing valid
inference from incomplete datasets. \textbf{add societal relevance: what
is the consequence of non-convergence?}

\textbf{Include sub-questions and add structure of the paper!} How can
non-convergence be identified diagnostically? Are common MCMC
non-convergence diagnostics appropriate for MICE? And if so, which
threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)? \textbf{How severe is it when the algorithm
has not converged? And what are guidelines for practice? Can the
parameter of interest, estimand \(Q\), be correct when the algorithm is
not (yet) converged, and vice versa?} \emph{(Maybe add these too? What
are the effects of continued iterations on estimates, predictions and
inferences? Do the answers differ with varying missingness proportions?
That is, we vary the nr of iterations and the missingness proportion
because we assume that the algorithm has not reached convergence at
\(t=1\), and performs worse with moderate to high missingness (not so
much with little or a LOT of missingness).)}

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on
\(k\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\)
(\(j = 1, 2, \dots, k\)) may be either observed or missing. \textbf{The
number of units \(i\) in dataset \(y\) with at least one missing data
value, divided by the total number of units \(n\), is called the
missingness proportion in \(p_{mis}\).} The collection of observed data
values in \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is
referred to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
\(M \times T\) times plausible values, where \(M\) is the number of
imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of
iterations (\(t = 1, 2, \dots, T\)). The collection of samples between
the initial value (at \(t=1\)) and the final imputed value (at \(t=T\))
will be referred to as an `imputation chain'. \textbf{Add: \(\theta\)s
are scalar summaries of interest in the iterative algorithm (e.g., chain
means; the average of the imputed values in each imputation chain). }

@@

\textbf{The goal of imputation through e.g.~MICE is to obtain a
multivariate distribution by iterating over the sequence of univariate
imputations. Traditionally, the inspection of the multivariate
convergence is done by visually inspecting the univariate chain means
and variances over the iterations, or by tracking the \(\widehat{R}\)
and autocorrelation. The downside to these approaches is that they
either focus on the univariate state space, or primarily track the
change over the iterations of a multivariate outcome conform the
scientific model of interest. Focusing on the convergence of these
outcome parameters may influence this procedure in the sense that the
model of evaluation favors the model of interest. Ideally, one would
like to evaluate a model-independent parameter that summarizes the
multivariate nature of the data. We propose eigenvalue as such a
parameter, because it\ldots. }

Mixing and stationarity have histroically been inspected visually, by
evaluating traceplots of scalar summaries of interest (\(\theta\)s;
e.g., chain means and chain variances). As \citet{buur18} describes,
users can also specify a model-specific scalar summary (e.g., a
regression coefficient). A user-specified scalar summary, however, is
not universal to all complete data problems. Therefore, as inspired by
\citep{mack03}, we propose a \(\theta\) that summarizes the multivariate
state space of the algorithm. Namely, the first eigenvalue of the
variance-covariance matrix of the \(M\) completed datasets. \textbf{The
first eigenvalue has the appealing property that is not dependent on the
substantive model of interest.}

\textbf{Since convergence of iterative imputation algorithms is in
distribution, there is not a unique point at which the algorithm reaches
a converged state. We can only evaluate scalar summaries of the
multivariate state space of the algorithm, \(\theta\)s. The recommended
scalar summaries to evaluate are chain means and chain variances.
Additionally, researchers may ``monitor some statistic of scientific
interest'' \citep[\(\S\) 6.5.2]{buur18}. Because of the issues mentioned
above, we propose a novel \(\theta\) to monitor: the first eigenvalue of
the variance-covariance matrix of the completed data. Let
\(\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j\) be the eigenvalues
of \(\Sigma\) in each of the \(M\) completed datasets
\({y_{obs}, y_{imp}}\). \(\lambda_1\) is measure that summarises the
covariances in the completed datasets.}

There are two requirements for convergence of iterative algorithms:
mixing and stationarity \citep{gelm13}. Without mixing, imputation
chains do not intermingle nicely, \textbf{indicating that \ldots{}}.
Without stationarity, there is trending within imputation chains, which
implies that further iterations would yiled a different set of
imputations.

To illustrate what non-convergence looks like in MI, we reproduce the
example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure
\ref{fig:non-conv} shows the traceplot for one of the variables in the
example. we see the average of the imputed values for a variable \(j\)
in \(y_{imp}\). The average imputed values in the left hand plot are
about a magnitude 2 lower than the right hand plot. So non-convergence
has a impact on the average imputed value for \(j\) in \(y_{imp, m}\).
This difference (presumably) translates to bias in the pooled estimate
\(\bar{\mu}_j\) as well. And maybe also to higher order statistics,
e.g.~variances, covariances, etc. Therefore, it's important to converge.
\textbf{Explain what we see, namely example by \citet{buur18}
reproduced, showing the traceplots of chain means for some variable. The
first plot is typical convergence of MICE, the second is pathological
non-convergence because of a mis-specified imputation model. Each line
is an imputation. In the first plot, the chains intermingle nicely and
there is little to no trending. In the second plot, there is a lot of
trending and some chains do not intermingle. Importantly, the chain
means at the last iteration (the imputed value per \(m\)) are very
different between the two plots. The algorithm with the mis-specified
model yields imputed values that are on average a magnitude two larger
than those of the typically converged algorithm. This shows the
importance of reaching converged states in iterative imputation
algorithms.}

\hypertarget{methods-under-evaluation}{%
\subsection{Methods under evaluation}\label{methods-under-evaluation}}

Non-stationarity within chains may be diagnosed with e.g.,
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, use we \(\widehat{R}\) and \(AC\) to evaluate
mixing and stationarity separately, as recommended by e.g.,
\citet{cowl96}.

@@

\caption{Typical convergence versus pathological non-convergence. Please note that this is the same data with a different imputation model, leading to different imputations (see the y axis!). Wel min of meer dezelfde origin op iteration 1.}\label{fig:non-conv}
\end{figure}

@@

An updated version of \(\widehat{R}\) has been proposed by
\citet{veht19} (p.~5). \textbf{This version may be suitable for
iterative imputation.} Let \(M\) be the total number of chains, \(T\)
the number of iterations per chain, and \(\theta\) the scalar summary of
interest (e.g., chain mean or chain variance). For each chain
(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
average these to obtain within-chain variance \(W\).

@@

We can interpret \(AC\)-values as a measure of stationarity. If
\(AC\)-values are close to zero, there is no dependence between
subsequent samples within imputation chains. Negative \(AC\)-values
indicate divergence within imputation chains. \textbf{Subsequent sampled
values within each imputation chain are less alike.} Positive
\(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent
iterations are similar, trending may occur. Negative \(AC\)-values show
no threat to the stationarity component of convergence. On the contrary
even---negative \(AC\)-values indicate that \(\theta\)-values of
subsequent iterations diverge from one another, which may increase the
variance of \(\theta\) and speed up convergence. As convergence
diagnostic, the interest is therefore in positive \(AC\)-values.
\textbf{Maybe remove:} Moreover, the magnitude of \(AC\)-values may be
evaluated statistically, but that is outside of our scope.

@@

Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
thresholds for MCMC algorithms, because of its convergence to a
distribution. In practice, non-convergence is usually diagnosed when
\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
performed to assess whether \(AC\) is significantly different from zero.

We assume that complete convergence, defined as \(\widehat{R} = 1\) and
\(AC = 0\), will not be observed. Because even in the most converged
state, the algorithm will show some signs of non-mixing and
non-stationarity. Upon \emph{sufficient} convergence, the imputation
chains will intermingle such that the only difference between the chains
is caused by the randomness induced by the algorithm
(\(\widehat{R} > 1\)), and there may be some dependency between
subsequent iterations of imputation chains (\(AC > 0\)).

\textbf{Before using them in the simulation, the two methods show at
least be able to distinguish between the two algorithms plotted in
Figure 2. From visual inspection we know that the typical convergence
has some signs of non-mixing around iteration 2, and little trending. In
the non-convergence situation, there is a lot of trending upto iteration
6, after which the chains reach a somewhat more stationary state. Mixing
in this situation gets worse from the first iteration onward, and
gradually gets better around iteration 7. To assess whether
\(\widehat{R}\) and \(AC\) may be appropriate non-convergence
identifiers for iterative imputation algorithms, the following condition
must hold: the methods indicate worse performance for the mis-specified
model with pathological non-convergence (i.e., higher \(\widehat{R}\)-
and \(AC\)-values than the typical performance). And additionally the
methods should reflect the increasing convergence in the typical
convergence situation as the number of iterations goes up.}

\textbf{In Figure \ref{fig:diagnostics}A, the chain means from Figure
\ref{fig:non-conv} are plotted again, now together---as are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up }in the typical convergence situation\textbf{. The
superior performance of the typical convergence over the pathological
non-convergence is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function
\texttt{stats::acf()}. When we look at this panel, we conclude something
weird. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for
the typical convergence and the pathological non-convergence, while
there is obvious trending in the latter. Moreover, the best convergence
(as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending up-to iteration number seven. After consulting the
documentation on stats::acf() \citep{R}, we conclude that this
implementation of \(AC\) is not suitable for iterative imputation
algorithms. There is a correction factor for a mathematical shortcut
that works in the limit. According to \citet{box15}, the function works
when the number of iterations \textgreater{} 50. The default number of
iterations in iterative imputation, however, is often much lower.
Therefore, we compute \(AC\) manually, see panel D. The \(AC\) values in
this plot do meet the requirements (\ldots) and will therefore be used
in the simulation study.}

\textbf{acf is niet bedoeld om weinig iteraties te\ldots{} niet een fout
in acf, alleen hier niet van toepassing}

@@

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence in the MICE algorithm on several quantities of
scientific interest \(Q\). Subsequently we will evaluate how well
\(\widehat{R}\) and \(AC\) perform in identifying the effects of
non-convergence (\textbf{add thetas}). And finally, we will formulate an
informed advice on the requirements to safely assume \emph{sufficient}
convergence in practice. That is, we assume that the algorithm is
\emph{sufficiently} converged when each estimate \(\bar{Q}\) is an
unbiased and confidence valid estimate of the corresponding estimand
\(Q\). (\textbf{add to formulate advice on thetas?})

@@

>>>>

Anyone who analyzes person-data may run into a missing data problem.
Missing data is not only ubiquitous, but treating it can also be
tedious. If a dataset contains just one incomplete observation,
statistical inferences are undefined and will not produce any results.
To circumvent this, many statistical packages employ list-wise deletion
by default (i.e., ignoring incomplete observations). Unfortunately, this
\emph{ad hoc} solution may yield wildly invalid results \citep{buur18}.
An alternative is to \emph{impute} (i.e., fill in) the missing values in
the incomplete observations. Subsequently, statistical inferences can be
performed on the completed dataset. By repeating this process several
times, a distribution of plausible results may be obtained, which
reflects the uncertainty in the data due to missingness. This technique
is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven
to be a powerful technique to draw valid inferences from incomplete data
under many circumstances \citep{buur18}.

Figure \ref{fig:diagram} provides an overview of the steps involved with
MI. Missing data \(y_{mis}\) in an incomplete dataset is imputed \(M\)
times. The imputed data \(y_{imp}\) is combined with the observed data
\(y_{obs}\) to create \(M\) completed datasets. On each completed
dataset, an analysis of scientific interest is performed. The quantity
of scientific interest (e.g., a regression coefficient) is denoted with
\(Q\). Since \(Q\) is estimated on each completed dataset, \(M\)
separate \(\hat{Q}\)-values are obtained. These \(M\) values are
combined into a single pooled estimate \(\bar{Q}\). The premise of
multiple imputation is that \(\bar{Q}\) is an unbiased and confidence
valid estimate of the true---but missing---data inference.

@@

\caption{Scheme of the main steps in multiple imputation (where $M=3$)---from an incomplete dataset, to $M$ multiply imputed datasets, to $M$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
\end{figure}

A popular method to obtain imputations is to use the `Multiple
Imputation by Chained Equations' algorithm, shorthand
`MICE'\citep{mice}. With MICE, imputed values \(y_{imp}\) are drawn from
the posterior predictive distribution of the missing values \(y_{mis}\).
The algorithm is named after the iterative algorithmic procedure by
which imputed values are generated: a multivariate distribution is
obtained by iterating over a sequence of univariate imputations. The
iterative nature of algorithms like MICE introduces a potential threat
to the validity of the imputations: What if the algorithm has not
converged? Are the implications then to be trusted? And can we rely on
the inference obtained on the completed data? These are all open
questions, because the convergence properties of iterative imputation
algorithms have not been systematically studied \citep{buur18}.
Moreover, there is no scientific consensus on how to evaluate
convergence of imputation algorithms \citep{taka17}. Some default MICE
techniques (e.g., `predictive mean modeling') might not yield converged
states at all \citep{murr18}. Therefore, algorithmic convergence should
be monitored carefully.

Currently, the recommended practice for evaulating the convergence of
iterative imputation algorithms is through visual inspection. That is,
the state space of the algorithm is monitored by plotting a scalar
summary \(\theta\) against the iteration number. \textbf{{[}Segway to
WHAT TO MONITOR? Also, this paragraph is more or less the same as the
third paragraph in the second section. Merge or remove one of the
two?{]}} Typically, the \(\theta\)s under evaluation are chain means and
variances (\textbf{define!}). These \(\theta\)s may be insufficient
because they are univariate summaries of the state space of the
algorithm, while MICE is not only concerned with univariate estimates,
but the entire multivariate distribution of \(y_{imp}\). A suggestion by
\citet{buur18} for a multivariate \(\theta\) to monitor is the quantity
of scientific interest \(Q\) (e.g., a regression coefficient).
Implementing this, however, might be somewhat advanced for empirical
researchers. Moreover, this scalar summary is not model-independent,
i.e., it only applies to one model of scientific interest. Yet, one of
the advantages of MI is that the missing data problem and scientific
problem are solved independently. Therefore, this \(\theta\) is not
sufficient either. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also
proposed multivariable evaluation of the MICE algorithm through
eigenvalue decomposition, building on the work of \citet{mack03}.
Eigenvalues of a variance-covariance matrix are a measure of the data's
total covariance. This would be a model-independent, multivariate scalar
summary to monitor, but it has not yet been implemented.
\textbf{{[}Segway to HOW TO MONITOR?{]}} Monitoring convergence visually
is insufficient on two counts: 1) it may be challenging to the untrained
eye, and 2) only severely pathological cases of non-convergence may be
diagnosed \citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative,
diagnostic evaluation of convergence would be preferred---although not
straightforward. Iterative imputation algorithms such as MICE are Markov
chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not
from a scalar to a point, but from one distribution to another. The
values generated by the algorithm (e.g., imputed values) will vary even
after convergence. Since MCMC algorithms do not reach a unique point at
which convergence is established, diagnostics may only identify signs of
\emph{non}-convergence \citep{hoff09}. Several of such non-convergence
diagnostics exist, but it is not known whether these are appropriate for
iterative imputation algorithms.

In this paper, we investigate how non-convergence in iterative
imputation algorithms may be diagnosed. First, we define which
non-convergence diagnostics will be considered, and evaluate how these
diagnostics could be appropriate for iterative imputation applications.
In the second part of this paper, we will investigate the impact of
inducing non-convergence in iterative imputation algorithms by means of
model-based simulation in \texttt{R} \citep{R}. The aim of the
simulation is to determine whether unbiased, confidence valid inferences
may be obtained under different levels of non-convergence (\textbf{and
at which point convergence may safely be assumed}). Additionally, we
will assess how well different non-convergence diagnostics perform in
identifying signs of non-convergence. The results of this study are
guidelines for identifying non-convergence in iterative imptation
algorithms, which will aid applied researchers in drawing valid
inference from incomplete datasets. \textbf{{[}add more societal
relevance: what is the consequence of non-convergence? e.g., biased
estimates, invalid inference, etc.{]}} For reasons of brevity, we only
focus on the iterative imputation algorithm implemented in the popular
\texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.

\textbf{{[}Include sub-questions and make structure of the paper
clear!{]}} How can non-convergence be identified diagnostically? Are
common MCMC non-convergence diagnostics appropriate for MICE? And if so,
which threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)? How severe is it when the algorithm has not
converged? And what are guidelines for practice? Can the parameter of
interest \(Q\) be correct when the algorithm is not (yet) converged, and
vice versa?

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on
\(k\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\)
(\(j = 1, 2, \dots, k\)) may be either observed or missing. The number
of units \(i\) in dataset \(y\) with at least one missing value, divided
by the total number of units \(n\), is called the missingness proportion
\(p_{mis}\). The collection of observed data values in \(y\) is denoted
by \(y_{obs}\); the missing part of \(y\) is referred to as \(y_{mis}\).
For each datapoint in \(y_{mis}\), we sample \(M \times T\) plausible
values, where \(M\) is the number of imputations
(\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations
(\(t = 1, 2, \dots, T\)) in the imputation algorithm. The collection of
samples between the initial value (at \(t=1\)) and the final imputed
value (at \(t=T\)) will be referred to as an `imputation chain'.
\textbf{\(\theta\)s are scalar summaries of the algorithm at a certain
iteration (e.g., chain means; the average of the imputed values in each
imputation chain). }

@@

There are two requirements for convergence of iterative algorithms:
mixing and stationarity \citep{gelm13}. Without mixing, imputation
chains do not intermingle nicely, indicating that the distribution of
imputed values differs across imputations. Chains may be `stuck' at a
local optimum, instead of sampling imputed values from the entire
predictive posterior distribution of the missing values. This may cause
under-estimation of the variance within chains, which results in
spurious, invalid inferences. Without stationarity, there is trending
within imputation chains. Trending implies that further iterations would
yield a different set of imputations. Iterative imputation algorithms
that have not yet reached stationarity, may thus yield biased estimates.

\textbf{{[}Move to INTRO???{]}} To illustrate what non-convergence looks
like in iterative imputation algorithms, we reproduce an example from
van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv}
displays a subset of the example: a variable that we will call \(j\). We
see two traceplots of the chain means for \(j\)---i.e., the average of
the imputed values for \(j\) in each imputation \(y_{imp,m}\), plotted
against the iteration number. Each line portrays one imputation. The
plot on the lefthand side of the figure shows typical convergence of an
iterative imputation algorithm. The righthand side displays pathological
non-convergence, induced by purposefully mis-specifying the imputation
model. \textbf{{[}Leave this in if it's after mixing and stationarity
are defined, otherwise remove?{]} In the typical convergence situation,
the imputation chains intermingle nicely and there is little to no
trending. In the non-convergence plot, there is a lot of trending and
some chains do not intermingle.} Importantly, the chain means at the
last iteration (the imputed value per \(m\)) are very different between
the two plots. The algorithm with the mis-specified model yields imputed
values that are on average a factor two larger than those of the
typically converged algorithm. That means that non-convergence has a
impact on the mean of \(j\) in \(y_{imp, m}\). This difference
(presumably) translates to bias in the \(M\) completed data estimates
\(\hat{\mu}_{j,m}\), and consequently also to the pooled estimate
\(\bar{\mu}_j\). Non-convergence thus leads to biased estimates. This
shows the importance of reaching converged states in iterative
imputation algorithms.

@@

\caption{Typical convergence versus pathological non-convergence. Please note that this is the same data with a different imputation model, leading to different imputations (e.g., see the units on the y-axis).}\label{fig:non-conv}
\end{figure}

\textbf{{[}Move to INTRO???{]}} Convergence has historically been
inspected visually, by monitoring the imputation algorithm over
iterations. This is typically done through traceplots. In a traceplot,
the iteraton number is plotted against a certain \(\theta\). \(\theta\)s
are scalar summaries of the state space of the algorithm at a specific
iteration. The recommended scalar summaries to evaluate are chain means
and chain variances. Additionally, researchers may ``monitor some
statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As
\citet{buur18} describes, researchers can specify their quantity of
scientific interest \(Q\) (e.g., a regression coefficient) as scalar
summary \(\theta\). Such a user-defined scalar summary, however, may be
somewhat advanced for empirical researchers. And moreover, it is not
universal to all complete data problems. Focusing on the convergence of
outcome parameters may influence the iterative imputation procedure in
the sense that the model of evaluation favors the model of interest. The
downside to these two approaches is that they either focus on the
univariate state space, or primarily track the change over the
iterations of a multivariate outcome conform the scientific model of
interest. Ideally, one would like to evaluate a model-independent
parameter that summarizes the multivariate nature of the data.
Therefore, we propose a \(\theta\) that summarizes the multivariate
state space of the algorithm, but is independent from the model of
scientific interest. We propose \(\lambda_1\) as such a scalar summary.
We define \(\lambda_1\) as the first eigenvalue of the
variance-covariance matrix of the completed data. Let
\(\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j\) be the eigenvalues
of \(\Sigma\) in each of the \(M\) completed datasets
\({y_{obs}, y_{imp}}\). \(\lambda_1\) is measure that summarises the
covariances in the completed datasets. \textbf{The first eigenvalue has
the appealing property that is not dependent on the substantive model of
interest. Eigenvalue decomposition is inspired by \citet{mack03}.}

\hypertarget{diagnostics-under-evaluation}{%
\subsection{Diagnostics under
evaluation}\label{diagnostics-under-evaluation}}

Non-convergence diagnostics for MCMC algorithms typically identify
problems in either one of the two requirements for converegnce: mixing
or stationarity. Non-stationarity within chains may be diagnosed with
e.g., autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric
standard error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, we use \(\widehat{R}\) and \(AC\) to evaluate
mixing and stationarity separately, as recommended by e.g.,
\citet{cowl96}.

@@

An updated version of \(\widehat{R}\) has been proposed by Vehtari et
al.~\citeyearpar[p.~5]{veht19}. \textbf{This version may be suitable for
iterative imputation.} Let \(M\) be the total number of chains, \(T\)
the number of iterations per chain, and \(\theta\) the scalar summary of
interest (e.g., chain means). For each chain (\(m = 1, 2, \dots, M\)),
we estimate the variance of \(\theta\), and average these to obtain
within-chain variance \(W\).

@@

We can interpret \(AC\)-values as a measure of stationarity. If
\(AC\)-values are close to zero, there is no dependence between
subsequent samples within imputation chains. Negative \(AC\)-values
indicate divergence within imputation chains. \textbf{Subsequent sampled
values within each imputation chain are less alike.} Positive
\(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent
iterations are similar, trending may occur. Negative \(AC\)-values show
no threat to the stationarity component of convergence. On the contrary
even---negative \(AC\)-values indicate that \(\theta\)-values of
subsequent iterations diverge from one another, which may increase the
variance of \(\theta\) and speed up convergence. As convergence
diagnostic, the interest is therefore in positive \(AC\)-values. The
magnitude of \(AC\)-values may be evaluated statistically {[}\textbf{add
reference}{]}.

\hypertarget{thresholds}{%
\subsubsection{Thresholds}\label{thresholds}}

Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
thresholds for MCMC algorithms, because of its convergence to a
distribution. Even in the most converged state, the algorithm will show
some signs of non-mixing and non-stationarity. Upon \emph{sufficient}
convergence, the imputation chains will intermingle such that the only
difference between the chains is caused by the randomness induced by the
algorithm (\(\widehat{R} > 1\)), and there may be some dependency
between subsequent iterations of imputation chains (\(AC > 0\)). In
practice, non-convergence is usually diagnosed when \(\widehat{R}\)
\textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is performed to
assess whether \(AC\) is significantly different from zero.

@@

Before we evaluate the performance of the non-convergence diagnostics
\(\widehat{R}=1\) and \(AC=0\) quantitatively through simulation, we
apply them to the example of pathological non-convergence by
\citet{buur18} that we reproduced above. We want to be able to draw the
same conclusions as we did from visually inspecting the two algorithms.
But \(\widehat{R}=1\) and \(AC=0\) should at least be able to
distinguish between the two algorithms plotted in Figure
\ref{fig:non-conv}. From visual inspection, we know that the typically
converged algorithm initially portrayed some signs of non-mixing (around
\(t=2\)), but intermigled nicely overall. Additionally, there was very
little trending. The algorithm with pathological non-convergence showed
severe non-mixing, although this gradually improved (from \(t=7\)).
Moreover, there was a lot of trending as well initially (up-to \(t=6\)),
after which the chains reached a somewhat more stationary state. To
assess whether \(\widehat{R}\) and \(AC\) may be appropriate
non-convergence identifiers for iterative imputation algorithms, they
should identify the same trends as the visual inspection. But at least,
the following condition must hold: the methods should indicate worse
performance for the mis-specified model with pathological
non-convergence than the typically converged algorithm (i.e., higher
\(\widehat{R}\)- and \(AC\)-values).

In Figure \ref{fig:diagnostics}A, the chain means from Figure
\ref{fig:non-conv} are plotted again---now together. Panel B shows two
versions of calculating \(AC\)s applied to the \(\theta\)s of panel A:
the default calculation with \texttt{R} function \texttt{stats::acf()}
\citep{R}, and manual calculation as the correlation between \(\theta\)
in iteration \(t\) and \(\theta\) in iteration \(t+1\). Panel C shows
the traditional computation of \(\widehat{R}\), and panel D shows
\(\widehat{R}\) as computed by implementing \citet{veht19} 's
recommendations.

@@

When we look at panel B, we conclude something weird. The \(AC\)-values
calculated with the default function indicate equal performance for the
typical convergence and the pathological non-convergence (up-to
\(t=5\)), while there is obvious trending in the \(\theta\)s of the
latter. Moreover, the best convergence (as indicated by the lowest
\(AC\)-value) is observed at \(t=2\), but looking at the chain means in
panel A, there should be some signs of trending up-to iteration number
seven. After consulting the documentation on \texttt{stats::acf()}
\citep{R}, we conclude that this \(AC\) function is not suited for
iterative imputation algorithms. The function is optimized for
performance when \(t\geq50\) \citep{box15}, while the default number of
iterations in iterative imputation is often much lower. Therefore, we
compute \(AC\) manually, see the solid line. These \(AC\) values
indicate non-stationarity as expected. We therefore calculate \(AC\)
manually.

\textbf{Describe panel C here and why it is not optimal{]}} The
\(\widehat{R}\)-values in panel D do meet the requirements specified
above. \(\widehat{R}\) as computed conform \citet{veht19} indeed
indicate less signs of non-convergence as the number of iterations goes
up.

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence in the MICE algorithm on several quantities of
scientific interest \(Q\). Subsequently we will evaluate how well
\(\widehat{R}\) and \(AC\) perform in identifying the effects of
non-convergence \textbf{{[}and performance of different thetas{]}}. And
finally, we will formulate an informed advice on the requirements to
safely assume \emph{sufficient} convergence in practice. That is, we
assume that the algorithm is \emph{sufficiently} converged when each
estimate \(\bar{Q}\) is an unbiased and confidence valid estimate of the
corresponding estimand \(Q\). \textbf{{[}and formulate advice on
thetas?{]}}

@@

Uitsmijter: To conclude, in the case of drawing inference from
incomplete data, convergence of the iterative imputation algorithm is
often a convenience but not a necessity.
